---
title: 'Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Lab 2'
geometry: margin=1in
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
---

```{r load packages, echo = FALSE, message = FALSE, warning=FALSE}
library(tidyverse)
library(tsibble)
library(latex2exp)
library(feasts)
library(patchwork)
library(forecast)
library(tseries)
library(knitr)
library(fable)
library(distributional)
#install.packages("kableExtra")
library(kableExtra)
#install.packages("imputeTS")
library(imputeTS)


# # Define a function to install and load a package if not already installed
# install_if_missing <- function(pkg) {
#   if (!requireNamespace(pkg, quietly = TRUE)) {
#     install.packages(pkg)
#   }
#   library(pkg, character.only = TRUE)
# }
# 
# # List of required packages
# packages <- c("tidyverse", "tsibble", "latex2exp", "feasts", "patchwork", 
#               "forecast", "tseries", "knitr", "fable", "distributional", 
#               "kableExtra", "imputeTS")
# 
# # Install and load each package
# invisible(lapply(packages, install_if_missing)) # hide the output

theme_set(theme_bw())
# theme_set(theme_minimal())
knitr::opts_chunk$set(dpi=1000)
```

\newpage
# Report from the Point of View of 1997
## (3 points) Task 0a: Introduction

Climate change is an increasingly pertinent issue for scientists and policymakers alike, as global temperatures rise. It is crucial to understand the underlying reasons for this increase, and its relationship with carbon emissions. This report presents potential outcomes of this constant increase, and highlights the need to anticipate future impacts of carbon emission reduction efforts.

Geochemist Dr. Charles David Keeling's pioneering work in atmospheric carbon dioxide measurements fundamentally reshaped our understanding of the global carbon cycle and its impact on climate change. In 1958, Keeling initiated a long-term study at the Mauna Loa Observatory, producing the iconic "Keeling Curve," which revealed the steady rise of atmospheric CO2. His research confirmed that fossil fuel combustion was contributing to increasing CO2 levels, a discovery with profound social and political consequences. This work also paved the way for further investigations into other greenhouse gases and established benchmarks for testing climate models.

CO2 is classified as a “greenhouse gas,” meaning that it traps heat in the atmosphere and lead to rising global temperatures when in high concentrations. It can be important to track Co2 levels as rising global temperatures can lead to imbalances in ecosystems and rising water levels that impact both animal and human life. Monitoring CO2 levels is critical because rising concentrations contribute to global warming, with severe consequences for ecosystems, sea levels, and both human and animal life. Understanding these trends is essential for assessing the long-term impact of human activities and guiding future policies.

## (3 points) Task 1a: CO2 data

The current data is gathered from measurements made under Dr. Charles Keeling's study at the Mauna Loa Observatory in Hawaii (Cleaveland, 1993). Measurements were taken by a chemical gas analyzer sensor, with detections based on infrared absorption. This data measures monthly CO2 concentration levels from January 1959 to December 1997. Units are in parts per million of CO2 (abbreviated as ppmv) using the SIO manometric mole fraction scale. Dr. Keeling initially designed a device to detect Co2 concentrations to detect CO2 emitted from limestone near bodies of water. But his measurements revealed a pattern of increasing CO2 concentrations at the global scale, urging further need to continue tracking the gas (Keeling, 1998). The time series shows a clear upward trend of global CO2 concentrations from 1959 to 1998, with an average increase in 1.26 CO2 ppmv and a standard deviation of .51 CO2 ppmv. Upon inspection of the yearly increases, the bulk of changing CO2 levels are between 0.5 and 2.0 CO2 ppmv.

```{r time-series-and-hist, warning=FALSE, echo = FALSE,fig.align='center', fig.height=4, fig.width=10, out.width="100%", fig.cap = "Data source: CO2 measurements from Mauna Loa Observatory"}
# import data
co2_tsib <- as_tsibble(co2)
# monthly time series
co2_trend_plot <- co2_tsib %>% 
  ggplot(aes(x = index, y = value)) +
  geom_line(color = 'black', size = .5) +
  labs(title = 'Increasing CO2 concentration from 1959 to 1997',x = 'Year', y = 'CO2 concentrations (ppmv)') +
  scale_x_yearmonth(date_breaks = "5 years")

# average_yearly_increase
co2_tsib_yearly_change <- co2_tsib %>% as_tibble() %>% 
  mutate(year = year(index)) %>% 
  group_by(year) %>% 
  summarise(`yearly_co2` = mean(value)) %>% 
    ungroup() %>% 
    mutate(lag_co2 = lag(yearly_co2),
           change = yearly_co2 - lag_co2, 
           percent_change = ((yearly_co2 - lag_co2)/lag_co2)*100)

# getting average increase (i.e. size of the trend)
yearly_mean <- mean(co2_tsib_yearly_change$change, na.rm = T) # average 1.26 units of co2 change each year
yearly_sd <- sd(co2_tsib_yearly_change$change, na.rm = T) # with a sd of .51 units of co2

change_hist <- ggplot(co2_tsib_yearly_change, aes(x = change)) + 
  geom_histogram(color = 'gray20', fill = 'gray', binwidth = .25) +
  scale_x_continuous(breaks = seq(floor(min(co2_tsib_yearly_change$change, na.rm =T)),
                                  ceiling(max(co2_tsib_yearly_change$change, na.rm =T)), by = 0.25)) +
  labs(
    title = "Histogram of Yearly Changes in CO2 ppmv",
    x = "Change",
    y = "Frequency"
  )

co2_trend_plot / change_hist

```

The time series also shows strong evidence of seasonality corresponding closely with the meteorological seasons of Autumn, Winter, Spring, and Summer. We now look at the ACF plot and average CO2 concentration for each month to gain further clarity on the seasonality.

```{r seasonality-eda, echo = FALSE,fig.align='center', fig.height=4, fig.width=10, out.width="100%",fig.cap = "Observing seasonality in CO2 concentration"}
# inspecting acf and graph of co2 concentrations over time

co2_acf <- acf(co2_tsib$value, plot = F)

co2_acf_plot <-  autoplot(co2_acf) + 
  labs(title = "ACF plot of monthly CO2 Concentrations", x = 'Lag', y = 'Autocorrelation')


monthly_co2_ave_plot <- co2_tsib %>% as_tibble() %>% 
  mutate(month = month(index)) %>% 
  group_by(month) %>% 
  summarise(co2_monthly_ave = mean(value, na.rm = T)) %>% 
  mutate(month_str = factor(month.abb[month], levels = month.abb)) %>% 
  ungroup() %>% 
  ggplot(aes(x = month_str, y = co2_monthly_ave, group = 1)) +
  geom_line(size = .8, color = 'purple4') +
  geom_point(size = 1.5) +
  ggtitle("Average CO2 Concentrations Across\nEach Month") +
  xlab('Month') +
  ylab('Co2 Concentrations (ppmv)') +
  theme(axis.text.x = element_text(angle = 45))


co2_acf_plot | monthly_co2_ave_plot
```
We also see a scallop/wave shaped pattern among correlations between the current value with growing lags. Clearer evidence of seasonality is shown when inspecting the monthly average the Co2 ppmv, when averaged across all years in the available data. CO2 contration peaks at the start of summer, and drops to a low in the fall, before rising again. This is likely due to the organic decomposition of plant life in these seasons (Keeling, 1960). 

We now study the time series' stationarity. We first conduct the Augmented Dickey-Fuller Test to test the null hypothesis that the time series is not stationary. As seen in the time series plot for `co2`, we have a clear upward trend, suggesting non-stationarity. This is confirmed by a p-value of 0.2269 yielded by the test, which indicates insufficient evidence to reject the null hypothesis of non-stationarity. To look at stationarity in variance, we fit a yearly CO2 average on the monthly time series, and inspect the residuals from year to year. Although there are slight changes in the variance, they seem to regress to a constant variance over time. Thus, once we account for the yearly increases in CO2 ppmv, there is likely a constant variance over time.

```{r seasonality-irregularities, echo = FALSE, fig.align='center', fig.height=3, fig.width=10, out.width="100%",fig.cap = "Studying variance over time"}

# checking for stationarity
#adf.test(co2_tsib$value)


# making a plot to show how the relationship looks like with yearly averages over the seasons
yearly_ave_w_residuals <- co2_tsib %>% as_tibble() %>% 
  mutate(year = year(index)) %>% 
  group_by(year) %>% 
  mutate(`Yearly Co2` = mean(value)) %>%
  mutate(residual = value - `Yearly Co2`) %>%
  pivot_longer(cols = c(value, `Yearly Co2`, residual), names_to = "type", values_to = "Monthly Co2") %>% 
  mutate(residual_bool = if_else(type == "residual", "Residuals", "Monthly Time Series Plotted on Yearly Average Co2")) 


yearly_ave_w_residuals_plot <- yearly_ave_w_residuals %>% 
  ggplot(aes(x = index, y = `Monthly Co2`, color = type)) +
  geom_line() +
  facet_wrap(~residual_bool, scales = "free_y", ncol = 1) +
  xlab('Date') +
  theme(legend.position = "none")

yearly_ave_w_residuals_plot

```


\newpage
## (3 points) Task 2a: Linear time trend model

We now fit a linear time trend model the `co2` series, and examine the characteristics of the fit and residuals.

```{r linear quad and poly models , echo = FALSE}
# Fit linear, quadratic, and cubic (3rd-degree polynomial) models at once
co2_models <- co2_tsib %>%
  model(
    linear_model = TSLM(value ~ index),           # Linear time trend model
    quad_model = TSLM(value ~ poly(index, 2)),    # Quadratic time trend model
    poly_model = TSLM(value ~ poly(index, 3))     # Cubic polynomial time trend model
  )

# Extract fitted values and residuals for all models
fitted_values <- co2_models %>% fitted()  # Fitted values for each model
residuals_values <- co2_models %>% residuals()  # Residuals for each model

# Add fitted values and residuals back to the tsibble
residuals_tslm <- residuals_values %>%
      pivot_wider(names_from = .model, values_from = .resid) %>%
    rename(
    linear_model_resid = linear_model,
    quad_model_resid = quad_model,
    poly_model_resid = poly_model
    )

fitted_tslm <- fitted_values %>%
      pivot_wider(names_from = .model, values_from = .fitted) %>%
    rename(
    linear_model_fitted = linear_model,
    quad_model_fitted = quad_model,
    poly_model_fitted = poly_model
    )

co2_tsib <- co2_tsib %>%
  left_join(residuals_tslm, by = "index") %>%
  left_join(fitted_tslm, by = "index")

```


```{r linear-model-plots,echo = FALSE,fig.align='center', fig.height=5, fig.width=10, out.width="100%",fig.cap = "Evaluating a Linear Model"}

# Plot time series with linear model line
lm_plot <- ggplot(co2_tsib, aes(x = index)) +
  geom_line(aes(y = value), color = 'black', size = .5) +  # Original time series
  geom_line(aes(y = linear_model_fitted), color = 'blue', size = .8) +  # Linear model predictions
  labs(title = 'CO2 Time Series with Linear Trend', x = 'Year', y = 'CO2 concentrations (ppmv)') +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

# Plot residuals with horizontal line at 0
lm_residuals_plot <- ggplot(co2_tsib, aes(x = index, y = linear_model_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at 0
  labs(title = "Residuals of Linear Time Trend Model", x = "Year", y = "Residuals") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

# Display the plots
lm_plot / lm_residuals_plot
```

Upon inspection of linear fit, the fitted line appears to be systematically overestimating values at certain points and underestimating values at other points. This indicates that perhaps a higher order polynomial might produce a better fit of the overall trend. The residuals of the linear model also exhibit a cyclical, non-linear pattern, indicating that the model does not capture the  seasonality in the data. The overall curve also suggests that the linear model insufficiently captures the overall trend. We now try a quadratic model, which may better capture the underlying trend.


```{r quadratic-model-plots,echo = FALSE,fig.align='center', fig.height=4.5, fig.width=10, out.width="100%",fig.cap = "Evaluating a Quadratic Model"}

# Plot time series with quadratic model predictions
quad_plot <- ggplot(co2_tsib, aes(x = index)) +
  geom_line(aes(y = value), color = 'black', size = .5) +  # Original time series
  geom_line(aes(y = quad_model_fitted), color = 'blue', size = .8) +  # Quadratic model predictions
  labs(title = 'CO2 Time Series with Quadratic Trend', x = 'Year', y = 'CO2 concentrations (ppmv)') +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

# Plot residuals with horizontal line at 0
quad_residuals_plot <- ggplot(co2_tsib, aes(x = index, y = quad_model_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at 0
  labs(title = "Residuals of Quadratic Time Trend Model", x = "Year", y = "Residuals") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

# Display the plots
quad_plot / quad_residuals_plot
```

\newpage
The quadratic model's residuals indicate a small reduction in variance, demonstrating a slightly improved fit. However, the cyclical behavior remains, indicating that seasonality is unaccounted for in the model still. There is also an overall non-random trend in the residuals, indicating that the model still may not capture all the structural details. We now fit a polynomial model to the data to see if there is an improved fit.

```{r polynomial-model-plots, echo = FALSE,fig.align='center', fig.height=5, fig.width=10, out.width="100%",fig.cap = "Evaluating a Polynomial (3) Model"}

# Plot time series with linear model line
poly_plot <- ggplot(co2_tsib, aes(x = index)) +
  geom_line(aes(y=value), color = 'black', size = .5) +  # Original time series
  geom_line(aes(y=poly_model_fitted), color = 'blue', size = .8) +  # Poly model predictions
  labs(title = 'CO2 Time Series with Polynomial (3) Trend', x = 'Year', y = 'CO2 concentrations (ppmv)') +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

# Plot residuals
poly_residuals_plot <- ggplot(co2_tsib, aes(x = index, y = poly_model_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at 0
  labs(title = "Residuals of Polynomial (3) Time Trend Model", x = "Year", y = "Residuals")

poly_plot/
poly_residuals_plot
```

The third-order polynomial model demonstrates improved residual behavior compared to quadratic and linear models. We chose to stop at this order to prevent excessive overfitting, as higher-order polynomials showed diminishing returns in model performance.

Apart from transforming the orders of the model, we were interested in data transformations - specifically logarithmic. As such we expermiented with a logarithmic dataset to observe the pattern of the data values.

```{r log-transformed-plot, echo = FALSE,fig.align='center', fig.height=3, fig.width=5, out.width="80%",fig.cap = "Evaluating effect of taking a log on the series"}
# Log transformation
co2_tsib$log_value <- log(co2_tsib$value)

# Plot log-transformed data
log_data_plot <- ggplot(co2_tsib, aes(x = index, y = log_value)) +
  geom_line() +
  labs(title = "Log-transformed Co2 Concentrations", x = "Year", y = "Log of Co2 concentrations")

log_data_plot
```

The logarithmic transformation reduces variance but offers minimal improvement compared to traditional plotting. This limited impact is likely due to the cyclical nature of the time series, which the transformation does not adequately address.

To address the cyclical behavior, we developed another polynomial model that includes each month as a variable. The average monthly CO2 emissions indicate significant cyclic patterns at the monthly level. By incorporating this variable, we anticipate an improvement in the fit of our time series model.

```{r seasonal polynomial data ,echo = FALSE}
# Create seasonal dummy variables
co2_tsib$month <- factor(month(co2_tsib$index))
co2_tsib$year <- factor(year(co2_tsib$index))

# Define a function to convert months into seasons
co2_tsib$season <- case_when(
  month(co2_tsib$index) %in% c(12, 1, 2) ~ "Winter",
  month(co2_tsib$index) %in% c(3, 4, 5) ~ "Spring",
  month(co2_tsib$index) %in% c(6, 7, 8) ~ "Summer",
  month(co2_tsib$index) %in% c(9, 10, 11) ~ "Autumn"
)

# Convert season into a factor
co2_tsib$season <- factor(co2_tsib$season, levels = c("Winter", "Spring", "Summer", "Autumn"))

# Fit poly (3) models with month variable and season variable
co2_models_x <- co2_tsib %>%
  model(
    poly_month_model = TSLM(value ~ poly(index, 3, raw = TRUE) + month),  # Polynomial (3rd degree) with month as factor
    poly_season_model = TSLM(value ~ poly(index, 3, raw = TRUE) + season) # Polynomial (3rd degree) with season as factor
  )

# Extract fitted values and residuals for all models
fitted_values_x <- co2_models_x %>% fitted()  # Fitted values for each model
residuals_values_x <- co2_models_x %>% residuals()  # Residuals for each model

# Add fitted values and residuals back to the tsibble
residuals_tslm_x <- residuals_values_x %>%
      pivot_wider(names_from = .model, values_from = .resid) %>%
    rename(
    poly_month_model_resid = poly_month_model,
    poly_season_model_resid = poly_season_model
    )

fitted_tslm_x <- fitted_values_x %>%
      pivot_wider(names_from = .model, values_from = .fitted) %>%
    rename(
    poly_month_model_fitted = poly_month_model,
    poly_season_model_fitted = poly_season_model
    )

co2_tsib <- co2_tsib %>%
  left_join(residuals_tslm_x, by = "index") %>%
  left_join(fitted_tslm_x, by = "index")

```


```{r plot-month, echo = FALSE, fig.align='center', fig.height=4, fig.width=11, out.width="100%",fig.cap = "Evaluating a Polynomial (3) with month variable model"}
# Plot time series with poly (3) and month variable
poly_month_plot <- ggplot(co2_tsib, aes(x = index)) +
  geom_line(aes(y=value), color = 'black', size = .5) +  # Original time series
  geom_line(aes(y=poly_month_model_fitted), color = 'blue', size = .8) +  # Poly model predictions
  labs(title = 'CO2 Time Series with Polynomial (3) and Month Variable', x = 'Year', y = 'CO2 concentrations (ppmv)') +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

# Plot residuals for poly (3) and month variable model
poly_month_residuals_plot <- ggplot(co2_tsib, aes(x = index, y = poly_month_model_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at 0
  labs(title = "Residuals of Polynomial (3) and Month Variable Model", x = "Year", y = "Residuals")

# Plot time series with poly (3) and season variable
poly_season_plot <- ggplot(co2_tsib, aes(x = index)) +
  geom_line(aes(y=value), color = 'black', size = .5) +  # Original time series
  geom_line(aes(y=poly_season_model_fitted), color = 'blue', size = .8) +  # Poly model predictions
  labs(title = 'CO2 Time Series with Polynomial (3) and Season Variable', x = 'Year', y = 'CO2 concentrations (ppmv)') +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

# Plot residuals for poly (3) and season variable model
poly_season_residuals_plot <- ggplot(co2_tsib, aes(x = index, y = poly_season_model_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at 0
  labs(title = "Residuals of Polynomial (3) and Season Variable Model", x = "Year", y = "Residuals")

poly_month_plot | poly_month_residuals_plot

```
Incorporating the `month` dummy variable brought the residuals closer to zero, ranging between 1 and -1, but they still displayed a seasonal pattern. To refine the model, we grouped the months into quarters, to represent the seasons as a categorical variable, `season`.

```{r plot-seasonal,  echo = FALSE, fig.align='center', fig.height=4, fig.width=11, out.width="100%",fig.cap = "Evaluating a Polynomial (3) with season variable model"}
poly_season_plot | poly_season_residuals_plot
```
We see that using the `season` variable centered the residuals around zero with a random distribution, though fluctuations remained between 2 and -2. We now proceed with this model, the polynomial model with the `season` dummy variable, as it has the residuals plot that most look like a random distribution around the red line centered on zero, and developed a forecast for CO2 emissions through 2020.
```{r forecasting-model, echo = FALSE,fig.align='center', fig.height=4, fig.width=10, out.width="100%",fig.cap = "Forecasting CO2 levels up to 2020 using a Polynomial (3) with month variable model"}
# Generate future monthly dates starting from January 1998 to December 2020
future_index <- seq(from = as.Date("1998-01-01"),  # Start in January 1998
                    to = as.Date("2020-12-01"), by = "1 month")  # Generate monthly dates up to Dec 2020

# Convert future_index to yearmonth format
future_index <- yearmonth(future_index)

# Create future data with month and season columns
future_data <- tibble(
  index = future_index,  # Future index (dates)
  month = factor(month(index)),
  year = factor(year(index)))

# Define a function to convert months into seasons
future_data$season <- case_when(
  month(future_data$index) %in% c(12, 1, 2) ~ "Winter",
  month(future_data$index) %in% c(3, 4, 5) ~ "Spring",
  month(future_data$index) %in% c(6, 7, 8) ~ "Summer",
  month(future_data$index) %in% c(9, 10, 11) ~ "Autumn"
)

# Convert season into a factor
future_data$season <- factor(future_data$season, levels = c("Winter", "Spring", "Summer", "Autumn"))

future_data<- future_data %>%
  as_tsibble(index = index)

forecast_poly_season <- forecast(co2_models_x$poly_season_model[[1]],
  new_data = future_data)  # Use a forecast horizon from the last observed point

autoplot(co2_tsib, value) +  # Plot original data  # Plot original data
  #geom_line(aes(x = index, y = poly_season_model_fitted), color = "blue", size = 1) +  # Add fitted values
  autolayer(forecast_poly_season, color = "blue2") +  # Add forecasted values
  labs(title = "CO2 Levels: Observed (1958-1997) and Forecasted (1998-2020)",
       x = "Year", y = "CO2 concentrations (ppmv)") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y") +
  guides(colour = guide_legend(title = "Series"))
```
The forecast model using the `season` variable shows decent performance, and predicts for the upward trend to persist up to 2020, along with the annual seasonality. We will now explore an ARIMA model to see if it may better capture the time series' underlying patterns and improve forecast accuracy.

## (3 points) Task 3a: ARIMA times series model

As seen in our EDA, there is non-stationarity. Thus, we will proceed to difference the data to make it stationary, both at the 1st lag followed by the 12th lag to account for seasonality, which is a crucial step before fitting the ARIMA model effectively.

```{r arima-differencing, warning=FALSE, echo = FALSE, fig.align='center', fig.height=3.5, fig.width=10, out.width="100%",fig.cap = "Differenced series looks more stationary in the mean and variance"}
# Apply both first-order and seasonal (12th-order) differencing
co2_tsib <- co2_tsib %>%
  mutate(diff_value = difference(difference(value, lag = 12),lag=1)) # First-order and seasonal differencing

# Remove leading NAs caused by differencing
co2_tsib <- co2_tsib %>% filter(!is.na(diff_value))

# Plot the differenced series to check if it looks stationary
diff_plot <- ggplot(co2_tsib, aes(x = index, y = diff_value)) +
  geom_line() +
  labs(title = "Differenced CO2 Concentrations", x = "Year", y = "Differenced CO2 concentrations (ppmv)") +
    scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y")

diff_plot 

# post-check for stationarity
# remove NA for adf test
# adf.test(na.omit(co2_tsib$diff_value))
```
The plot of the differenced time series does look more stationary in the mean and variance; which is confirmed by a 0.01 p-value yielded from the Augmented Dickey-Fuller Test, indicating sufficient evidence to reject the null hypothesis of non-stationarity. We now look at the ACF and PACF plots of the differenced time series, to inform how we should construct our ARIMA model.

\newpage

```{r acf-pacf,warning=FALSE,echo=FALSE,fig.align='center',fig.height=4,fig.width=10,out.width="100%",fig.cap = "Evaluating ACF and PACF for differenced series"}
acf_diff <- ACF(co2_tsib, diff_value, lag_max = 24) %>%
  autoplot() +
  labs(
    title = "ACF for Differenced CO2 Concentration",
    x = "Year",
    y = "Differenced CO2 Concentrations (ppmv)"
  )

pacf_diff <- PACF(co2_tsib, diff_value, lag_max = 24) %>%
  autoplot() +
  labs(
    title = "PACF for Differenced CO2 Concentration",
    x = "Year",
    y = "Differenced CO2 Concentrations (ppmv)"
  )

acf_diff | pacf_diff
```
Both ACF and PACF plots show strong auto-correlation with lag 1, and the ACF cutting off strong after lag 1, and the PACF having a significant spike at lag 1, and tapering off a little more. This might indicate that our model has a MA(1) component. The spike at lag 12 in the ACF might also indicate a seasonal MA component. 

```{r model bic, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE}
# Fit ARIMA model by testing different lags using the BIC criterion
model.bic <- co2_tsib %>%
  model(
    auto = ARIMA(value, stepwise = FALSE, approx = FALSE, ic = "bic")
    )

# Extract the report from the best ARIMA model
# model.bic %>%
#   select(auto) %>%
#   report()
```
The ARIMA function behaved as expected, and returned an ARIMA(0,1,1)(0,1,1)$_{12}$ function, with BIC = 182.32. We will now look at the residuals for this model.

```{r arima-residuals, fig.align='center',fig.height=4.5,fig.width=12,out.width="100%",message = FALSE, echo = FALSE, warning = FALSE,fig.cap = "Evaluating ARIMA model "}
# Extract residuals from the ARIMA model
residuals_arima <- residuals(model.bic)

# Plot the residuals over time
residual_plot <- autoplot(residuals_arima) +
  labs(title = "Residuals of ARIMA(0,1,1)(0,1,1)[12]", x = "Year", y = "Residuals")

# Plot the ACF of the residuals
acf_plot <- ggAcf(residuals_arima) +
  labs(title = "ACF of ARIMA(0,1,1)(0,1,1)[12] Residuals", x = "Lag", y = "ACF")

# Q-Q plot to check normality of residuals
qq_plot <- ggplot(data = as.data.frame(residuals_arima), aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of ARIMA(0,1,1)(0,1,1)[12] Residuals", x = "Theoretical Quantiles", y = "Sample Quantiles")

residual_plot|acf_plot |qq_plot


```

```{r ljung box test on residuals, echo = FALSE, warning= FALSE}
# Ljung Box Test on residuals
resid.ts<-model.bic %>%
  augment() %>%
  dplyr::select(.resid) %>%    # dplyr makes sure this doesn't bug out
  as.ts()

# Box.test(resid.ts, lag = 24, type = "Ljung-Box")
```
The residuals look random, there are no significant autocorrelations in the ACF, and they closely follow a a normal distribution in the Q-Q plot. These all indicate that the model has captured most of the underlying structure of our time series. A Ljung-Box test also yielded a p-value of 0.6733, further confirming that there is insufficient statistical evidence to reject the null hypothesis that there is no autocorrelation. We can proceed with forecasting the time series data through 2022.
\newpage

```{r arima-forecasting, fig.align='center',fig.height=4,fig.width=10,out.width="100%", echo = FALSE, warning= FALSE,fig.cap = "Forecasting up to 2022 using an ARIMA model", fig.pos = "H"}

# Generate forecasts from model.bic
forecast_arima <- model.bic %>%
  forecast(h = "25 years")  # Forecast horizon up to 2022

# Plot the forecasts along with the observed data
autoplot(co2_tsib, value) +  # Plot original data
  autolayer(forecast_arima, color = "blue2") +  # Add forecast 
  labs(title = "CO2 Levels: Observed and Forecasted (up to 2022)",
       x = "Year", y = "CO2 concentrations (ppmv)") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y") +
  guides(colour = guide_legend(title = "Series"))
```

## (3 points) Task 4a: Forecast atmospheric CO2 growth

We now forecast when atmospheric CO2 is expected to be at 420 ppm and 500 ppm for the first and final times, as seen in the table below. 

```{r long term forecast part one, echo = FALSE}
# Generate forecasts for a long horizon (up to 2100)
long_term_forecast <- model.bic %>%
  forecast(h = "103 years")  # Forecast horizon (assuming we want predictions up to 2100)

# Convert the forecast to a data frame for easier inspection
forecast_df <- as.data.frame(long_term_forecast)

# Look at when CO2 levels are predicted to reach 420 ppm
threshold_420 <- forecast_df %>%
  filter(.mean >= 420 & .mean < 421)

# Extract the mean and variance from threshold_420
mean_420 <- threshold_420$.mean  # Mean value
variance_420 <- variance(threshold_420$value)  # Extract variance from the distribution

# Calculate standard deviation
sd_420 <- sqrt(variance_420)

# Calculate the 80% confidence interval
ci_lower_420 <- mean_420 - 1.28 * sd_420
ci_upper_420 <- mean_420 + 1.28 * sd_420

# Look at when CO2 levels are predicted to reach 500 ppm
threshold_500 <- forecast_df %>%
  filter(.mean >= 500 & .mean < 501)

# Extract the mean and variance from threshold_420
mean_500 <- threshold_500$.mean  # Mean value
variance_500 <- variance(threshold_500$value)  # Extract variance from the distribution

# Calculate standard deviation
sd_500 <- sqrt(variance_500)

# Calculate the 80% confidence interval
ci_lower_500 <- mean_500 - 1.28 * sd_500
ci_upper_500 <- mean_500 + 1.28 * sd_500

# Looking at confidence intervals for first and last time 420 level is reached
# c(ci_lower_420[1],  ci_upper_420[1])
# c(ci_lower_420[length(ci_lower_420)], ci_upper_420[length(ci_upper_420)])

# Looking at confidence intervals for first and last time 500 level is reached
# c(ci_lower_500[1],  ci_upper_500[1])
# c(ci_lower_500[length(ci_lower_500)], ci_upper_500[length(ci_upper_500)])


```


```{r forecastone, echo = FALSE, message = FALSE}
# Create a data frame with blanks
co2_table <- data.frame(
  `CO2 Level` = c("420 ppm", "500 ppm"),  # CO2 levels to track
  `First Month` = c("2031 May", "2083 Apr"),      
  `First Value` = c("420.1 (402.3, 438.0)", "500.4 (437.9,562.9)"),       
  `Last Month` = c("2035 Oct", "2085 Dec"),      
  `Last Value` = c("420.4 (399.5, 441.3)", "500.9 (435.7,566.2)")      
)

# Print the table using kable
kable(co2_table, caption = "CO2 Levels and Forecasted Times with 80% Confidence Intervals")%>%
  kable_styling(font_size =9,latex_options = "HOLD_position")  # Adjust the font size here (change to any desired value)
```

Our model also forecasts CO2 levels in the year 2100. Although these forecasts include a standard deviation, but these do not take into account the existing efforts to reduce global greenhouse gases, such as **insert efforts here**. Thus, since these are very human-activty dependent, it is unlikely to be super accurate.

```{r forecasttwo, echo = FALSE, message = FALSE}
# Extract the forecast for the year 2100
forecast_2100 <- forecast_df %>%
  filter(year(index) == 2100)

# Create variance column
forecast_2100$std <- sqrt(variance(forecast_2100$value))
forecast_2100_table<-forecast_2100 %>%
  dplyr::select("index",".mean","std") %>%
  mutate(.mean = round(.mean,1), std = round(std,1)) %>%
  rename(
    Date =index,
    Value = .mean,
    SD = std
  )
# Print forecast for 2100
kable(forecast_2100_table, caption = "CO2 Forecasts in 2100")%>%
  kable_styling(font_size = 9,latex_options = "HOLD_position")  # Adjust the font size here (change to any desired value)
```



\newpage
# Report from the Point of View of the Present

## (1 point) Task 0b: Introduction
Following our initial evaluation using Keeling’s data, we now seek to re-examine the original study to identify potential deviations in CO2 level predictions. Specifically, we aim to discern whether any discrepancies between predicted and actual CO2 levels since 1997 arise from the inherent limitations of prior models or from changes within the CO2-generating system itself.

Before 1997, CO2 data was collected using a chemical gas analyzer that relied on infrared absorption to measure monthly CO2 concentrations from January 1959 through December 1997. The present data collection approach, however, leverages a newer CO2 analyzer installed at Mauna Loa, employing Cavity Ring-Down Spectroscopy (CRDS). CRDS determines CO2 concentration by measuring the rate at which light is absorbed in an optical cavity, rather than the intensity. This method offers significant advantages, as it eliminates dependencies on light intensity and sample path length, providing absolute, highly accurate concentration measurements without frequent recalibration. Furthermore, CRDS is capable of hourly measurements, and its rigorous gas-flushing system ensures each sample’s reliability. These improvements in measurement precision and frequency offer researchers enhanced insight into CO2 trends, facilitating a more robust evaluation of the models' predictive capabilities and any potential systemic changes.

## (3 points) Task 1b: Create a modern data pipeline for Mona Loa CO2 data.

We created a data pipeline, pulling the most recent weekly data from the [Global Monitoring Laboratory](https://gml.noaa.gov/ccgg/trends/data.html) page. We now conduct a similar EDA for the updated time series data from Dr. Xin Lan's study of CO2 atmospheric trends.

```{r data_pipeline, echo = FALSE}
# get co2 data from Mauna Loa weekly CO2 records url
url <- "https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_weekly_mlo.txt"

# convert txt file to readable data
data_raw <- read.delim(url, skip = 37, header = FALSE, sep = "", 
                       stringsAsFactors = FALSE)

# Adjust these column names to match the file's structure exactly
colnames(data_raw) <- c("year", "month", "day", "date_decimal", "value", "ndays", 
                        "1_year_ago", "10_years_ago", "increase_since_1800")

# Add week numbering index
data_raw <- data_raw %>%
  mutate(index = row_number())

#  Clean 1_year_ago and 10_years_ago columns
data_processed <- data_raw %>%
  mutate(
    `value` = ifelse(`value` <= -999.99, NA, `value`),
    `1_year_ago` = ifelse(`1_year_ago` <= -999.99, NA, `1_year_ago`),
    `10_years_ago` = ifelse(`10_years_ago` <= -999.99, NA, `10_years_ago`)
  ) %>%
  mutate(datetime = make_datetime(year, month, day), # Convert the timestamp columns (year, month, day, hour) to a datetime object
         datetime = as.Date(datetime)) %>%
  filter(!is.na(value))  # Remove any NA values in CO2 concentration

# Ensure unique, ordered datetime values
data_clean <- data_processed %>%
  arrange(datetime) %>%
  distinct(datetime, .keep_all = TRUE) %>%  # Remove duplicates, if any
  dplyr::select(datetime, value)

# create time series object
co2_present <- ts(data_clean$value, frequency = 52, start = c(year(min(data_processed$datetime)), 1))
```

We first plot the time series data for CO2 concentrations, and a histogram of the yearly changes. 


```{r time-series-and-hist-two, warning=FALSE, echo = FALSE,fig.align='center', fig.height=4, fig.width=10, out.width="100%", fig.cap = "Data source: CO2 measurements from Mauna Loa Observatory"}
# import data
new_co2_tsib <- as_tsibble(data_processed, index = datetime)
# new_co2_tsib <- new_co2_tsib %>% mutate(datetime = yearweek(datetime))

# monthly time series with the line of best fit
new_co2_trend_plot <- new_co2_tsib %>% 
  ggplot(aes(x = datetime, y = value)) +
  geom_line(color = 'black', size = .5) +
  labs(title = 'Increasing CO2 concentration from 1997 to Present',x = 'Year', y = 'CO2 concentrations (ppmv)') +
  scale_x_yearmonth(date_breaks = "5 years")


# average_yearly_increase
new_co2_tsib_yearly_change <- new_co2_tsib %>% as_tibble() %>% 
  group_by(year) %>% 
  summarise(`yearly_co2` = mean(value)) %>% 
  ungroup() %>% 
  mutate(lag_co2 = lag(yearly_co2),
         change = yearly_co2 - lag_co2, 
         percent_change = ((yearly_co2 - lag_co2)/lag_co2)*100)

# getting average increase (i.e. size of the trend)
new_yearly_mean <- mean(new_co2_tsib_yearly_change$change, na.rm = T) # average 1.90 units of co2 change each year
new_yearly_sd <- sd(new_co2_tsib_yearly_change$change, na.rm = T) # with a sd of .61 units of co2

new_change_hist <- ggplot(new_co2_tsib_yearly_change, aes(x = change)) + 
  geom_histogram(color = 'gray20', fill = 'gray', binwidth = .25) +
  scale_x_continuous(breaks = seq(floor(min(new_co2_tsib_yearly_change$change, na.rm =T)),
                                  ceiling(max(new_co2_tsib_yearly_change$change, na.rm =T)), by = 0.25)) +
  labs(
    title = "Histogram of Yearly Changes in CO2 ppmv",
    x = "Change",
    y = "Frequency"
  )

new_co2_trend_plot / new_change_hist

```
From 1998 to 2024, the time series continues to show a clear upward trend of global CO2 concentrations, with an average increase in 1.90 CO2 ppmv and a standard deviation of .61 CO2 ppmv. These are larger values than those calculated in 1997, indicating that the time series had a larger average increase and standard deviation from 1998 to present day 2024. The histogram also shows a rightward shift in the distribution, indicating that the annual changes have increased.

We now look at the ACF plot and weekly CO2 concentration across every year to gain further clarity on the seasonality.
\newpage

```{r seasonality-eda-two, echo = FALSE,fig.align='center', fig.height=4, fig.width=11, out.width="100%",fig.cap = "Observing seasonality in CO2 concentration",warning = FALSE}
# inspecting acf and graph of co2 concentrations over time
new_co2_acf <- acf(new_co2_tsib$value, plot = F)
new_co2_acf_plot <-  autoplot(new_co2_acf) + 
  labs(title = "ACF plot of monthly CO2 Concentrations", x = 'Lag', y = 'Autocorrelation',subtitle="1997-Sep 2024")

# Monthly aggregated dataset
new_co2_monthly_averages <- data_clean %>%
  mutate(month = month(datetime)) %>%  # Convert datetime to month format
  group_by(month) %>%  # Group by month to aggregate monthly averages
  summarise(co2_monthly_ave = mean(value, na.rm = TRUE)) %>%  # Calculate monthly average
  ungroup() %>%
  as_tsibble(index = month)  # Convert to tsibble

# Convert your data to a tsibble (if not already in tsibble format)
data_tsibble <- as_tsibble(data_clean, index = datetime)
data_tsibble <- data_tsibble %>%
  fill_gaps()

# Use gg_season to plot seasonality
gg_season_plot <- gg_season(data_tsibble, value) +
  labs(
    title = "Seasonal Plot of Weekly CO2 Concentrations",
    x = "Week of the Year",
    y = "CO2 Concentrations (ppmv)"
  )

new_co2_acf_plot | gg_season_plot
```

We see strong persistent autocorrelation for many lags, indicating an overall trend. The time series also shows strong evidence of seasonality corresponding closely with the meteorological seasons: CO2 concentration peaks at the start of summer, and drops to a low in the fall, before rising again. These observations are consistent with the report from 1997.

We now study the time series' stationarity. We conduct the Augmented Dickey-Fuller Test to test the null hypothesis that the time series is not stationary. As seen in the time series plot for `co2`, we have a clear upward trend, suggesting non-stationarity. However, this is *contrasted* by a p-value of 0.01 yielded by the test, which indicates evidence to reject the null hypothesis of non-stationarity. To look at stationarity in variance, we fit a yearly CO2 average on the monthly time series, and inspect the residuals from year to year. Although there are slight changes in the variance, they seem to regress to a constant variance over time. Thus, once we account for the yearly increases in CO2 ppmv, there is likely a constant variance over time.

```{r seasonality-irregularities-two, echo = FALSE, fig.align='center', fig.height=3, fig.width=10, out.width="100%",fig.cap = "Studying variance over time"}

# checking for stationarity
#adf.test(co2_present)

# making a plot to show how the relationship looks like with yearly averages over the seasons
new_yearly_ave_w_residuals <- new_co2_tsib %>% as_tibble() %>% 
  mutate(year = year(datetime)) %>% 
  group_by(year) %>% 
  mutate(`Yearly_Co2` = mean(value)) %>%
  mutate(residual = value - `Yearly_Co2`) %>%
  pivot_longer(cols = c(value, `Yearly_Co2`, residual), names_to = "type", values_to = "Yearly Co2") %>% 
  mutate(residual_bool = if_else(type == "residual", "Residuals", "Monthly Time Series Plotted on Yearly Average Co2")) 


new_yearly_ave_w_residuals_plot <- new_yearly_ave_w_residuals %>% 
  ggplot(aes(x = datetime, y = `Yearly Co2`, color = type)) +
  geom_line() +
  facet_wrap(~residual_bool, scales = "free_y", ncol = 1) +
  xlab('Date') +
  theme(legend.position = "none")


new_yearly_ave_w_residuals_plot

```
The new data collection method and techniques yield similar trends in average CO2 levels across months and years, with no notable deviations from prior patterns. Additionally, the residuals display consistent patterns, indicating that the updated methods align closely with previous data quality and trend behavior. The only detected change was the results of the adf test which found the CO2 values to be stationary.

\newpage
## (1 point) Task 2b: Compare linear model forecasts against realized CO2

```{r comparing-to-forecast, echo = FALSE, fig.align='center', fig.height=3, fig.width=10, out.width="100%", fig.cap = "Forecasted with a Polynomial (3) with Season Variable Model"}

# Generate future monthly dates starting from January 1998 to September 2024
future_index <- seq(from = as.Date("1998-01-01"),  # Start in January 1998
                    to = as.Date("2024-09-30"), by = "1 month")  # Generate monthly dates up to Sep 2024

# Convert future_index to yearmonth format
future_index <- yearmonth(future_index)

# Create future data with month and season columns
future_data <- tibble(
  index = future_index,  # Future index (dates)
  month = factor(month(index)),
  year = factor(year(index)))

# Define a function to convert months into seasons
future_data$season <- case_when(
  month(future_data$index) %in% c(12, 1, 2) ~ "Winter",
  month(future_data$index) %in% c(3, 4, 5) ~ "Spring",
  month(future_data$index) %in% c(6, 7, 8) ~ "Summer",
  month(future_data$index) %in% c(9, 10, 11) ~ "Autumn"
)

# Convert season into a factor
future_data$season <- factor(future_data$season, levels = c("Winter", "Spring", "Summer", "Autumn"))

future_data<- future_data %>%
  as_tsibble(index = index)

forecast_poly_season <- co2_models_x %>%
  dplyr::select(poly_season_model) %>%
  forecast(new_data = future_data)  # Use a forecast horizon from the last observed point


autoplot(new_co2_tsib, value) +  # Plot original data  # Plot original data
  autolayer(forecast_poly_season, color = "blue2") +  # Add forecasted values
  labs(title = "Forecasted CO2 Concentrations Compared to Observed Values",
       x = "Year", y = "CO2 concentrations (ppmv)") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y") +
  guides(colour = guide_legend(title = "Series"))


```
When plotting the observed Co2 concentrations to the forecasted values from data up to 1997, we see a clear mismatch in predictions. The model had predicted a declining trend of Co2 concentrations, but actual concentrations increased fairly linearly. The declining trend was modeled with an expected curvilinear relationship with Co2 concentrations over time, as this fit better when looking at the time period between the 1950s and 1990s. But the actual time series may not be appropriately measured as a strict polynomial relationship, possibly depending on external causes such as human-lead efforts to reduce carbon emissions or efforts to restrict policy against reducing carbon emissions.

## (1 point) Task 3b: Compare ARIMA models forecasts against realized CO2

```{r comparing-to-arima, echo = FALSE, fig.align='center', fig.height=3, fig.width=10, out.width="100%",fig.cap="Forecasted with a ARIMA Model"}

# Generate forecasts from model.bic
forecast_arima <- model.bic %>%
  forecast(new_data = future_data)  # Forecast horizon up to 2024

autoplot(forecast_arima, color = "blue2") +  # Add forecast first
  autolayer(new_co2_tsib, value, color = "black") +  # Plot observed data on top
  labs(title = "Forecasted CO2 Concentrations Compared to Observed Values",
       x = "Year", y = "CO2 concentrations (ppmv)") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y") +
  guides(colour = guide_legend(title = "Series"))

  
```

The ARIMA model, while closer in trend to the observed values than the curvilinear model, still underestimates the increases in global Co2 concentrations. The yearly fluctuations match better than the previous forecast and in the short-term the forecast performs well, but within 5 years the forecast diverges from actual concentrations and by 2020 we see concentrations of ~12.5 Co2 ppmv higher than anticipated. By 2024, the observed values at barely at the tail end of the 95% confidence interval, and look unlikely to stay within the confidence interval should the forecast continue.

## (3 points) Task 4b: Evaluate the performance of 1997 linear and ARIMA models

Before, we expected that we would reach 420 ppmv around May of 2031, but the first date we past this threshold was in March of 2022. In other words, our predictions were nearly a decade off courses. This may imply a more concerning trajectory for negative environmental outcomes than previously anticipated.

We now use the weekly data to generate a month-average series from 1997 to present-day 2024, and compare the overall forecasting performance of both the Polynomial (3) with Season variable model, and the ARIMA model.


```{r comparing-forecasting-performance, echo = FALSE, fig.align='center', fig.height=3, fig.width=10, out.width="100%"}

# getting root mean absolute error and root mean squared errors to compare models

# getting actual values
actual_values <- new_co2_tsib %>% 
  index_by(month = yearmonth(datetime)) %>% 
  summarise(value = mean(value, na.rm = T)) %>% 
  filter(month >= yearmonth('1998 Jan') & month <= yearmonth('2024 Sep')) %>% 
  pull(value)

# poly forecasted values
poly_forecast_values <- forecast_poly_season %>% pull(.mean)
  
#arima forecasted values
arima_forecast_values <- forecast_arima %>% 
  filter(index <= yearmonth('2024 Sep')) %>% 
  pull(.mean)

# Mean Absolute Error (MAE)
mae_poly <- mean(abs(actual_values - poly_forecast_values)) # ~ 12.3
mae_arima <- mean(abs(actual_values - arima_forecast_values)) # ~ 5.8

# Root Mean Squared Error (RMSE)
rmse_poly <- sqrt(mean((actual_values - poly_forecast_values)^2)) # ~ 15.3
rmse_arima <- sqrt(mean((actual_values - arima_forecast_values)^2)) # ~ 7.1

# Compare all four values
# mae_poly
# mae_arima
# rmse_poly
# rmse_arima

# statistically comparing performance using the Diebold-Mariano test

# Calculate forecast errors
errors_poly <- actual_values - poly_forecast_values
errors_arima <- actual_values - arima_forecast_values

# Perform Diebold-Mariano test
# dm.test(errors_poly, errors_arima, alternative = "two.sided", h = 1, power = 2)
# p-value less than .05, may reject the null hypothesis that the two errors are equal

```
We do this by first comparing the Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) of each forecasting model, and conduct a significance test to formally compare the differences. We see that the polynomial season model has an MAE of ~ 15.8, while the ARIMA model has lower MAE of ~ 7.3. Similarly, we see that the polynomial season model has a higher RMSE of ~ 19.9, while the ARIMA model has lower RMSE of ~ 8.9. That the ARIMA model has lower errors suggests that it is a better fitting forecasting model. Finally, the conducted Diebold-Mariano Test, which tests for differences in forecast errors, returns a significant p-value of less than 2.2e-16, allowing us to reject the null hypothesis that the two forecasts have the same forecast accuracy.


## (4 points) Task 5b: Train best models on present data

We now seasonally adjust the weekly data, and split both seasonally-adjusted (SA) and non-seasonally-adjusted (NSA) series into training and test sets, using the last two years (2022-2024) of observations as the test sets. We also use spline interpolation for 18 missing values. We obtain the decomposition in the plots below.

```{r stl-decomposition, echo = FALSE, fig.align='center', fig.height=6, fig.width=10, out.width="100%", fig.cap = "Clear seasonality and trend seen from decomposition", message = FALSE, warning = FALSE}
new_co2_tsib <- new_co2_tsib %>%
  fill_gaps()

# Impute using spline interpolation for the 18 missing values
new_co2_tsib_impute <- new_co2_tsib %>%
  mutate(value = na_interpolation(value, option = "spline")) %>% # Using spline interpolation
  select(datetime, value)

# Use decomposition
stl_model <- new_co2_tsib_impute %>%
  model(STL(value ~ season(window = 52)))  # STL decomposition with seasonal component

# Extract the components (trend, season, remainder) from the STL model
stl_components <- components(stl_model)

# Plot the components to see how STL decomposed the data
stl_plot <- autoplot(stl_components) + 
  labs(title = "STL Decomposition of CO2 Data", 
       y = "CO2 concentrations (ppmv)")
stl_plot

```

```{r splitting train test set, echo = FALSE}
# Convert datetime to yearweek format
stl_components <- stl_components %>%
  mutate(datetime = yearweek(datetime), seasonally_adjusted_value = value - season_year)# Subtract the seasonal component

new_co2_tsib_impute <- new_co2_tsib_impute %>%
  mutate(datetime = yearweek(datetime))

# Define training and test date ranges
train_end_date <- yearweek("2022-09-25")  # End of September 2022
test_start_date <- yearweek("2022-10-02")  # Start of October 2022
test_end_date <- yearweek("2024-09-29")    # End of September 2024

# Seasonally Adjusted (SA) training and test sets
# SA Training set 
train_SA <- stl_components %>%
  dplyr::select(datetime, seasonally_adjusted_value) %>%
  filter(datetime <= train_end_date)

# SA Test set
test_SA <- stl_components %>%
  dplyr::select(datetime, seasonally_adjusted_value) %>%
  filter(datetime >= test_start_date & datetime <= test_end_date) 

# Non-Seasonally Adjusted (NSA) training and test sets
# NSA Training set 
train_NSA <- new_co2_tsib_impute %>%
  dplyr::select(datetime, value) %>%
  filter(datetime <= train_end_date)

# NSA Test set 
test_NSA <- new_co2_tsib_impute %>%
  dplyr::select(datetime, value) %>%
  filter(datetime >= test_start_date & datetime <= test_end_date) 
```

We now look at the ACFs and PACFs of the SA and NSA training data, with a difference of 1 to account of the overall trend, to consider what ARIMA models we can fit to it. 

\newpage

```{r acf-and-pacfs, echo = FALSE, fig.align='center', fig.height=5, fig.width=10, out.width="100%", fig.cap = "EDA on training and test sets"}
# Compute the ACF plot for NSA data
train_NSA_acf <- ACF(train_NSA, difference(difference(value,lag=1),lag=52), lag_max = 104) %>% autoplot()+ 
  labs(title = "ACF plot of NSA Weekly CO2 Concentrations",subtitle = "Training Set, differenced 1st and 52nd", x = 'Lag', y = 'Autocorrelation')

# Compute the PACF plot for NSA data
train_NSA_pacf <- PACF(train_NSA, difference(difference(value,lag=1),lag=52), lag_max = 104) %>% autoplot()+ 
  labs(title = "PACF plot of NSA Weekly CO2 Concentrations",subtitle = "Training Set, differenced 1st and 52nd", 
       x = 'Lag', y = 'Partial Autocorrelation')


# Compute the ACF plot for SA data
train_SA_acf <- ACF(train_SA, difference(difference(seasonally_adjusted_value,lag=1),lag=52), lag_max = 104) %>%
  autoplot() + 
  labs(title = "ACF plot of SA Weekly CO2 Concentrations", subtitle = "Training Set, differenced 1st and 52nd", 
       x = 'Lag', y = 'Partial Autocorrelation')


# Compute the PACF plot for SA data
train_SA_pacf <- PACF(train_SA, difference(difference(seasonally_adjusted_value,lag=1),lag=52), lag_max = 104) %>%
  autoplot() + 
  labs(title = "PACF plot of SA Weekly CO2 Concentrations", subtitle = "Training Set, differenced 1st and 52nd", 
       x = 'Lag', y = 'Partial Autocorrelation')

# Display the plots
(train_NSA_acf | train_NSA_pacf) / (train_SA_acf | train_SA_pacf)

```

We note that the NSA ACF shows a sharp tapering off, while the PACF tapers off more slowly. This indicates perhaps a MA model might be more suitable. We also still see spikes at teh 52th lags, indicating we could perhaps add a MA seasonal component, and also difference. We thus expect perhaps something like  ARIMA(0,1,1)(0,1,1); however, we will let the `ARIMA` function choose a model for us. For the SA data, since it was already de-trended and we also see a tapering ACF and sharply dropping PACF, we will simply let the `ARIMA` function pick the best model. We get the tables below.

```{r training models, echo = FALSE, cache = TRUE}
# NSA
# Fit ARIMA model by testing different lags using the BIC criterion
model.bic.present.nsa <- train_NSA %>%
  model(
    auto = ARIMA(value, stepwise = FALSE, ic = "bic")
    )

# Extract the glance
nsa.auto.report <- model.bic.present.nsa %>%
  dplyr::select(auto) %>%
  glance()

# Look at report for auto
# model.bic.present.nsa %>%
#   select(auto) %>%
#   report()

#SA
# Fit ARIMA model by testing different lags using the BIC criterion
model.bic.present.sa <- train_SA %>%
  model(
    #arima = ARIMA(seasonally_adjusted_value ~ 0 + pdq(0,1,4)))
    auto = ARIMA(seasonally_adjusted_value, stepwise = FALSE, ic = "bic"))
    

# Look at report for auto
# model.bic.present.sa %>%
#   select(auto) %>%
#   report()

# Extract the report from the best ARIMA model
sa.auto.report <- model.bic.present.sa %>%
  select(auto)%>%
  glance()

# NSA: Create a data frame for NSA models
nsa_table <- data.frame(
  Model = c("Auto: ARIMA(0,1,3)(2,1,0)[52]"),
  AICc = c(nsa.auto.report$AICc),
  BIC = c(nsa.auto.report$BIC),
  LogLik = c(nsa.auto.report$log_lik)
)

# SA: Create a data frame for SA model
sa_table <- data.frame(
  Model = c("Auto: ARIMA(4,1,1)"),
  AICc = c(sa.auto.report$AICc),
  BIC = c(sa.auto.report$BIC),
  LogLik = c(sa.auto.report$log_lik)
)

```

```{r nsa-table, echo = FALSE}
kable(nsa_table, caption = "NSA ARIMA Model Results")%>%
  kable_styling(font_size = 9,latex_options = "HOLD_position") 
```


```{r sa-table,echo = FALSE}
kable(sa_table, caption = "SA ARIMA Model Results")%>%
  kable_styling(font_size = 9,latex_options = "HOLD_position") 
```

Now we assess the residuals for the NSA model ARIMA(0,1,3)(2,1,0)[52]. We get the following plots.

```{r assessing-residuals-nsa, fig.align='center',fig.height=4.5,fig.width=12,out.width="100%",message = FALSE, echo = FALSE, warning = FALSE,fig.cap = "Evaluating ARIMA model for NSA data"}
# Extract residuals from the ARIMA model
residuals_arima.nsa <- model.bic.present.nsa %>%
  dplyr::select(auto) %>%
  residuals()
residuals_arima.nsa <- residuals_arima.nsa %>%
  mutate(datetime = as.Date(datetime))

# Plot the residuals over time
residual_plot.nsa <- autoplot(residuals_arima.nsa) +
  labs(title = "Residuals of ARIMA(0,1,3)(2,1,0)[52] ", x = "Year", y = "Residuals")

# Plot the ACF of the residuals
acf_plot.nsa <- ggAcf(residuals_arima.nsa) +
  labs(title = "ACF of ARIMA(0,1,3)(2,1,0)[52]  Residuals", x = "Lag", y = "ACF")

# Q-Q plot to check normality of residuals
qq_plot.nsa <- ggplot(data = as.data.frame(residuals_arima.nsa), aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of ARIMA(0,1,3)(2,1,0)[52]  Residuals", x = "Theoretical Quantiles", y = "Sample Quantiles")

residual_plot.nsa|acf_plot.nsa |qq_plot.nsa

```
They mostly look like white noise, and the q-q plot is moderately close to a normal distribution. A Ljung Box Test on the residuals yielded a p-value of 0.1173, indicating insufficient evidence to reject the null-hypothesis that there is no auto-correlation. We are good with this model, and now proceed to assess the residuals for the ACF plot. 
```{r ljung box test nsa, echo = FALSE, warning= FALSE}
# Ljung Box Test on residuals
resid.ts.nsa <- residuals_arima.nsa %>%
  dplyr::select(.resid) %>%
  pull(.resid)  # Convert to a numeric vector

# Convert the residuals into a time series object
resid.ts.nsa <- ts(resid.ts.nsa)

# Perform the Ljung-Box test on the residuals
# Box.test(resid.ts.nsa, lag = 52, type = "Ljung-Box")
```

```{r assessing-residuals-sa, fig.align='center',fig.height=4.5,fig.width=12,out.width="100%",message = FALSE, echo = FALSE, warning = FALSE,fig.cap = "Evaluating ARIMA model for SA data"}
# Extract residuals from the ARIMA model
residuals_arima.sa <- model.bic.present.sa %>%
  dplyr::select(auto) %>%
  residuals()
residuals_arima.sa <- residuals_arima.sa %>%
  mutate(datetime = as.Date(datetime))

# Plot the residuals over time
residual_plot.sa <- autoplot(residuals_arima.sa) +
  labs(title = "Residuals of ARIMA(4,1,1)", x = "Year", y = "Residuals")

# Plot the ACF of the residuals
acf_plot.sa <- ggAcf(residuals_arima.sa) +
  labs(title = "ACF of ARIMA(4,1,1) Residuals", x = "Lag", y = "ACF")

# Q-Q plot to check normality of residuals
qq_plot.sa <- ggplot(data = as.data.frame(residuals_arima.sa), aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of ARIMA(4,1,1) Residuals", x = "Theoretical Quantiles", y = "Sample Quantiles")

residual_plot.sa|acf_plot.sa |qq_plot.sa

```

```{r ljung box test sa, echo = FALSE, warning= FALSE}
# Ljung Box Test on residuals
resid.ts.sa <- residuals_arima.sa %>%
  dplyr::select(.resid) %>%
  pull(.resid)  # Convert to a numeric vector

# Convert the residuals into a time series object
resid.ts.sa <- ts(resid.ts.sa)

# Perform the Ljung-Box test on the residuals
#Box.test(resid.ts.sa, lag = 52, type = "Ljung-Box")
```
We can see that the residuals do not look like white noise, and a Ljung-Box test yielded a p-value of < 1.07e-09, indicating sufficient evidence to reject the null-hypothesis that there is no auto-correlation. We tried other models like an ARIMA(0,1,4) from studying the ACF and PACF plots, but it did not improve the results. Nonetheless, we now look at the forecasts of these two models on our test data. We now look at both the fit of our models on the actual data, as well as the forecasts for both models. 

\newpage


```{r final-forecasts, fig.align='center', fig.height=7, fig.width=10, out.width="100%", echo = FALSE, fig.cap = "Assessing ARIMA Models on NSA and SA data"}

# Extract fitted values for the NSA ARIMA model
nsa_fitted <- model.bic.present.nsa %>%
  dplyr::select(auto) %>%
  fitted()

# Generate forecasts for the NSA ARIMA model
nsa_forecast <- model.bic.present.nsa %>%
  dplyr::select(auto) %>%
  forecast(new_data = test_NSA)

# Plot the fitted values for training, forecast for test data, and actual data
nsa_forecast_plot<-autoplot(nsa_fitted,.fitted, color = "blue2") +  # Fitted values for training data
  autolayer(train_NSA, value, color = "black") +  # Actual values for training data
  autolayer(nsa_forecast, color = "blue2") +  # Forecast for test data
  autolayer(test_NSA, value, color = "black") +  # Actual values for test data
  labs(title = "NSA: Assessing ARIMA(1,1,1)(0,1,1)\nForecasted Values On Test Set", x = "Year", y = "CO2 concentrations (ppmv)") +
  coord_cartesian(xlim = c(as.Date("2020-01-01"), NA), ylim=c(375,450)) # Limit to data from 2020 onwards

nsa_full_plot<-autoplot(nsa_fitted,.fitted, color = "blue2") +  # Fitted values for training data
  autolayer(train_NSA, value, color = "black",alpha = 0.8) +  # Actual values for training data
  autolayer(nsa_forecast, color = "blue2") +  # Forecast for test data
  autolayer(test_NSA, value, color = "black") +  # Actual values for test data
  labs(title = "NSA: Comparing ARIMA(1,1,1)(0,1,1)\nFit on Training Data", x = "Year", y = "CO2 concentrations (ppmv)")

# Extract fitted values for the SA ARIMA model
sa_fitted <- model.bic.present.sa %>%
  dplyr::select(auto) %>%
  fitted()

# Generate forecasts for the SA ARIMA model
sa_forecast <- model.bic.present.sa %>%
  dplyr::select(auto) %>%
  forecast(new_data = test_SA)

# Plot the fitted values for training, forecast for test data, and actual data
sa_forecast_plot<-autoplot(sa_fitted,.fitted,color = "blue2") +  # Fitted values for training data
  autolayer(train_SA, seasonally_adjusted_value, color = "black") +  # Actual values for training data
  autolayer(sa_forecast,color = "blue2") +  # Forecast for test data
  autolayer(test_SA, seasonally_adjusted_value, color = "black") +  # Actual values for test data
  labs(title = "SA: Assessing ARIMA(1,1,0)\nForecasted Values On Test Set", x = "Year", y = "CO2 concentrations (ppmv)") +
  coord_cartesian(xlim = c(as.Date("2020-01-01"), NA), ylim=c(375,450))  # Limit to data from 2020 onwards

# Plot the fitted values for training, forecast for test data, and actual data
sa_full_plot<-autoplot(sa_fitted,.fitted,color = "blue2") +  # Fitted values for training data
  autolayer(train_SA, seasonally_adjusted_value, color = "black", alpha = 0.8) +  # Actual values for training data
  autolayer(sa_forecast,color = "blue2") +  # Forecast for test data
  autolayer(test_SA, seasonally_adjusted_value, color = "black") +  # Actual values for test data
  labs(title = "SA: Comparing ARIMA(1,1,0)\nFit on Training Data", x = "Year", y = "CO2 concentrations (ppmv)")
  
  
(sa_full_plot | nsa_full_plot)/
(sa_forecast_plot| nsa_forecast_plot)

```
We can see that our models' fitted values in blue are extremely close to the actual values plotted in black in the top two plots. For the forecasts, both models 95% confidence intervals in blue included the actual data in black However, the NSA model seemed to provide a closer fit to the actual data. We now compare both the RMSE and MAE of both models.

```{r MSE and MAE for final models, echo = FALSE}
# NSA: Calculate MSE and MAE for training set (fitted values)
nsa_train_errors <- nsa_fitted %>%
  left_join(train_NSA, by = "datetime") %>%
  mutate(
    se = (value - .fitted)^2,
    ae = abs(value - .fitted)
  )

nsa_rmse_train <- sqrt(mean(nsa_train_errors$se))
nsa_mae_train <- mean(nsa_train_errors$ae)

# SA: Calculate MSE and MAE for training set (fitted values)
sa_train_errors <- sa_fitted %>%
  left_join(train_SA, by = "datetime") %>%
  mutate(
    se = (seasonally_adjusted_value - .fitted)^2,
    ae = abs(seasonally_adjusted_value - .fitted)
  )
sa_rmse_train <- sqrt(mean(sa_train_errors$se))
sa_mae_train <- mean(sa_train_errors$ae)


# NSA: Calculate MSE and MAE for test set (forecasted values)
nsa_test_errors <- nsa_forecast %>%
  left_join(test_NSA, by = "datetime") %>%
  mutate(
    se = (value.y - .mean)^2,
    ae = abs(value.y - .mean)
  )

nsa_rmse_test <- sqrt(mean(nsa_test_errors$se))
nsa_mae_test <- mean(nsa_test_errors$ae)

# SA: Calculate MSE and MAE for test set (forecasted values)
sa_test_errors <- sa_forecast %>%
  left_join(test_SA, by = "datetime") %>%
  mutate(
    se = (seasonally_adjusted_value.y - .mean)^2,
    ae = abs(seasonally_adjusted_value.y - .mean)
  )
sa_rmse_test <- sqrt(mean(sa_test_errors$se))
sa_mae_test<- mean(sa_test_errors$ae)
```


```{r final-models-comparison-table, echo = FALSE}
# Create a data frame to store the RMSE and MAE results
model_comparison <- data.frame(
  Metric = c("RMSE (Training)", "MAE (Training)", "RMSE (Test)", "MAE (Test)"),
  `NSA ARIMA Model` = c(nsa_rmse_train, nsa_mae_train, nsa_rmse_test, nsa_mae_test),
  `SA ARIMA Model` = c(sa_rmse_train, sa_mae_train, sa_rmse_test, sa_mae_test)
)

# Use kable to display the comparison table
kable(model_comparison, caption = "Comparison of NSA and SA Models on Training and Test Sets")%>%
  kable_styling(font_size = 9,latex_options = "HOLD_position") 
```

As we see from the table above, the SA model outperformed the NSA model on the training set in both RMSE and MAE metrics, but the NSA model outperformed the SA model on the test set, indicating that the NSA model of ARIMA(0,1,3)(2,1,0) generalizes better. Given that our SA ARIMA model can be improved, we now fit a polynomial time-trend model for the SA series and compare it with the SA ARIMA model.

\newpage

```{r SA-poly-model, fig.align='center', fig.height=3.5, fig.width=10, out.width="100%", fig.cap = "Assessing Polynomia (3) Model on SA data", echo = FALSE}
SA_poly_model <- train_SA %>%
  model(
    poly_model = TSLM(seasonally_adjusted_value ~ poly(datetime, 3,raw = TRUE))     # Cubic polynomial time trend model
  )
fitted_SA_poly<- SA_poly_model %>%
  dplyr::select(poly_model) %>%
  fitted()

forecast_SA_poly <- SA_poly_model %>%
  dplyr::select(poly_model) %>%
  forecast(new_data = test_SA)  # Use a forecast horizon from the last observed point

SA_poly_full_plot <- autoplot(train_SA, seasonally_adjusted_value) +  # Plot original data  # Plot original data
  autolayer(fitted_SA_poly, .fitted, color = "blue2")+
  autolayer(forecast_SA_poly, color = "blue2") +  # Add forecasted values
  autolayer(test_SA, seasonally_adjusted_value)+
  labs(title = "Polynomial (3) Model Fit on SA Training Set",
       x = "Year", y = "CO2 concentrations (ppmv)") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y") +
  guides(colour = guide_legend(title = "Series"))


SA_poly_forecast_plot <- autoplot(train_SA, seasonally_adjusted_value) +  # Plot original data  # Plot original data
  autolayer(fitted_SA_poly, .fitted, color = "blue2")+
  autolayer(forecast_SA_poly, color = "blue2") +  # Add forecasted values
  autolayer(test_SA, seasonally_adjusted_value)+
  labs(title = "Assessing Polynomial (3) Model on SA Test Set",
       x = "Year", y = "CO2 concentrations (ppmv)") +
  scale_x_yearmonth(date_breaks = "5 years", date_labels = "%Y") +
  coord_cartesian(xlim = c(as.Date("2020-01-01"), NA), ylim=c(375,450)) # Limit to data from 2020 onwards

SA_poly_full_plot | SA_poly_forecast_plot

```
We can see that Polynomial (3) model seems to fit the overall trend decently, although it does not perform so well on the forecast, with the actual data out of the 95% confidence interval. The Polynomial (3) model also yielded a RMSE and MAE of 0.7558659 and 0.6160157 respectively for the training set, and a RMSE and MAE of 1.611581 and 1.377034 respectively for the test set. Although the ARIMA models outperformed the Polynomial (3) model by far on the training sets, the Polynomial (3) model did better with a lower RMSE and MAE on the test set.

```{r assesing poly three model, echo = FALSE}
# Calculate RMSE and MAE for poly (3) model
sa_poly_train_errors <- train_SA %>%
  left_join(fitted_SA_poly, by = "datetime") %>%
  mutate(
    se = (seasonally_adjusted_value - .fitted)^2,
    ae = abs(seasonally_adjusted_value - .fitted)
  )

sa_poly_test_errors <- test_SA %>%
  left_join(forecast_SA_poly, by = "datetime") %>%
  mutate(
    se = (seasonally_adjusted_value.x - .mean)^2,
    ae = abs(seasonally_adjusted_value.x - .mean)
  )

sa_poly_rmse_train <- sqrt(mean(sa_poly_train_errors$se))
sa__poly_mae_train<- mean(sa_poly_train_errors$ae)

sa_poly_rmse_test <- sqrt(mean(sa_poly_test_errors$se))
sa__poly_mae_test<- mean(sa_poly_test_errors$ae)

# Look at the values
# sa_poly_rmse_train
# sa__poly_mae_train
# sa_poly_rmse_test
# sa__poly_mae_test
```

## (3 points) Task Part 6b: How bad could it get?

With the non-seasonally adjusted data series, we now generate predictions for when atmospheric CO2 is expected to be at 420 ppm and 500 ppm levels for the first and final times. We also generate a prediction for atmospheric CO2 levels in the year 2122.

```{r future_prediction, echo = FALSE}
# Generate forecasts for the NSA ARIMA model
nsa_forecast <- model.bic.present.nsa %>%
  dplyr::select(auto) %>%
  forecast(h = "101 years") # up to 2122

# Convert the forecast to a data frame for easier inspection
forecast_df <- as.data.frame(nsa_forecast) %>%
  mutate(monthyear = yearmonth(datetime))

# Look at when CO2 levels are predicted to reach 420 ppm
threshold_420 <- forecast_df %>%
  filter(.mean >= 420 & .mean < 421)

# Extract the mean and variance from threshold_420
mean_420 <- threshold_420$.mean  # Mean value
variance_420 <- variance(threshold_420$value)  # Extract variance from the distribution

# Calculate standard deviation
sd_420 <- sqrt(variance_420)

# Calculate the 80% confidence interval
ci_lower_420 <- mean_420 - 1.28 * sd_420
ci_upper_420 <- mean_420 + 1.28 * sd_420

# Look at when CO2 levels are predicted to reach 500 ppm
threshold_500 <- forecast_df %>%
  filter(.mean >= 500 & .mean < 501)

# Extract the mean and variance from threshold_420
mean_500 <- threshold_500$.mean  # Mean value
variance_500 <- variance(threshold_500$value)  # Extract variance from the distribution

# Calculate standard deviation
sd_500 <- sqrt(variance_500)

# Calculate the 80% confidence interval
ci_lower_500 <- mean_500 - 1.28 * sd_500
ci_upper_500 <- mean_500 + 1.28 * sd_500

```

```{r match_values}
# Define a function to get the closest value
get_closest <- function(target, values) {
  values[which.min(abs(values - target))]
}

# get first and last values
last_420 <- c(ci_lower_420[length(ci_lower_420)], ci_upper_420[length(ci_upper_420)])
last_500 <- c(ci_lower_500[length(ci_lower_500)], ci_upper_500[length(ci_upper_500)])

# Find the closest .mean values in forecast_df to each target in last_420
closest_420 <- sapply(last_420, get_closest, values = forecast_df$.mean)

# Filter forecast_df for the closest .mean values for 420 ppm
matching_dates_420 <- forecast_df %>%
  filter(.mean %in% closest_420) %>%
  dplyr::select(monthyear, .mean)

# Find the closest .mean values in forecast_df to each target in last_500
closest_500 <- sapply(last_500, get_closest, values = forecast_df$.mean)

# Filter forecast_df for the closest .mean values for 500 ppm
matching_dates_500 <- forecast_df %>%
  filter(.mean %in% closest_500) %>%
  dplyr::select(monthyear, .mean)

# # Display the results
matching_dates_420
matching_dates_500

# Looking at confidence intervals for first and last time 420 and 500 level is reached
c(ci_lower_420[1],  ci_upper_420[1])
c(ci_lower_500[1],  ci_upper_500[1])

```


```{r forecast_420_500, echo = FALSE, message = FALSE}
# Create a data frame with blanks
co2_table <- data.frame(
  `CO2 Level` = c("420 ppm", "500 ppm"),  # CO2 levels to track
  `First Month` = c("2022 Nov", "2043 Jun"),      
  `First Value` = c("419.01", "468.6"),       
  `Last Month` = c("2024 Jul", "2068 Dec"),      
  `Last Value` = c("422.7", "532.9")      
)

# Print the table using kable
kable(co2_table, caption = "CO2 Levels and Forecasted Times ARIMA NSA with 80% Confidence Intervals")%>%
  kable_styling(font_size =9,latex_options = "HOLD_position")  # Adjust the font size here (change to any desired value)
```

Our model also forecasts CO2 levels in the year 2122. Although these forecasts include a standard deviation, but these do not take into account the existing efforts to reduce global greenhouse gases, such as the global effort to achieve net-zero GHG emissions by transitioning away from gas-cars to EVs. Thus, since these are very human-activity dependent, it is unlikely to be super accurate.

```{r forecast_2122, echo = FALSE, message = FALSE}
# Extract the forecast for the year 2100
forecast_2122 <- forecast_df %>%
  filter(year(datetime) == 2122)

# Create variance column
forecast_2122$std <- sqrt(variance(forecast_2122$value))
forecast_2122_table<-forecast_2122 %>%
  dplyr::select("monthyear",".mean","std") %>%
  mutate(.mean = round(.mean,1), std = round(std,1)) %>%
  rename(
    Date =monthyear,
    Value = .mean,
    SD = std
  )
# Print forecast for 2122
kable(forecast_2122_table, caption = "CO2 Forecasts in 2122")%>%
  kable_styling(font_size = 9,latex_options = "HOLD_position")  # Adjust the font size here (change to any desired value)
```

